Informing policy via dynamic models: Cholera in Haiti

### 2024-02-14 edits.tex
Added paragraph about fitting meta-pop stochastic model

>>>>

\newcommand\editMetaPop{In our literature review, 17 articles considered dynamic models that incorporate spatial heterogeneity \cite{lee20,tuite11,pasetto18,fitzgibbon20,eisenberg13,rinaldo12,chao11,abrams13,trevisin22,sallah17,collins14,kelly16,azman12,leung22,kuhn14,mari15,gatto12}. All but four \cite{lee20,pasetto18,sallah17,azman12} of these studies used deterministic dynamic models, which greatly simplifies the process of calibrating model parameters to incidence data. The model in \cite{pasetto18} was fit using an Ensemble Kalman Filter (EnKF) \cite{evensen09}; though EnKF scales favorably with the number of spatial units, it relies on linearization of latent states which can be problematic for highly nonlinear systems \cite{evensen22,ionides21}. Alternative approaches used to fit stochastic models included making additional simplifying assumptions to aid in the fitting process \cite{lee20}, and using MCMC algorithms \cite{sallah17,azman12} which require specific structures in the latent dynamics, making these algorithms non plug-and-play. In this subsection, we present how the recently developed iterated block particle filter (IBPF) algorithm \cite{ning21ibpf,ionides22} can be used to fit a spatially explicit stochastic dynamic model to incidence data.}

### 2024-02-14 ms.Rnw
Save conditional unit log-likelihoods obtained with the block-particle filter.
Added paragraph about fitting meta-pop stochastic model

@@

\jwc{This doesn't quite feel like the right place for this either, but I can't seem to pin down exactly where that would be yet. Perhaps the discussion?}
\new{\editMechModels}

@@

\subsubsection*{Calibrating Model~3 Parameters}\label{sec:fit3}
Model~3 describes cholera dynamics in Haiti using a metapopulation model, where the hidden states in each administrative department has an effect on the dynamics in other departments.
The decision to address metapopulation dynamics using a spatially explicit model, rather than to aggregate over space, is double-edged.
Evidence for the former approach has been provided in previous studies \cite{king15}, including the specific case of heterogeneity between Haitian departments in cholera transmission \cite{collins14}.
However, a legitimate preference for simplicity can support a decision to consider nationally aggregated models \cite{saltelli20,green15}.
One issue that arises when fitting spatially explicit models is that parameter estimation techniques based on the particle filter become computationally intractable as the number of spatial units increases.
This is a result of the approximation error of particle filters growing exponentially in the dimension of the model \cite{rebeschini15,park20}.

To avoid the approximation error present in high-dimensional models, Lee et al.~\cite{lee20} simplified the problem of estimating the parameters of Model~3 by creating an approximate version of the model where the units are independent given the observed data.
Reducing a spatially coupled model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
Because the simplified, spatially-decoupled version of Model~3 implemented in \cite{lee20} relies on the observed cholera cases, the calibrated model cannot readily be used to obtain forecasts.
Therefore, in order to obtain model forecasts, Lee et al.~\cite{lee20} used the parameters estimates from the spatially-decoupled approximation of Model~3 to obtain forecasts using the fully coupled version of the model.
This approach of model calibration and forecasting avoids the issue of particle depletion, but may also be problematic.
One concern is that cholera dynamics in department $u$ are highly related to the dynamics in the remaining departments;
calibrating model parameters while conditioning on the observed cases in other departments may therefore lead to an over-dependence on observed cholera cases.
Another concern is that the two versions of the model are not the same, resulting in sub-optimal parameter estimates for the spatially coupled model,
as parameters that maximize the likelihood of the decoupled model almost certainly do not maximize the likelihood of the fully coupled model.
These two concerns may explain the unrealistic forecasts and low likelihood of Model~3 of Lee et al.~\cite{lee20} (see Table~\ref{tab:likes}).

At the time Lee et al.~\cite{lee20} conducted their study, there was no known algorithm that could readily be used to maximize the likelihood of an arbitrary meta-population POMP model with coupled spatial dynamics, which justifies the spatial decoupling approximation that was used to calibrate model parameters.
For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the iterated block particle filter (IBPF) algorithm \cite{ionides22}.
This algorithm extends the work of Ning and Ionides \cite{ning21ibpf}, who provided theoretic justification for the version of the algorithm that only estimates unit-specific parameters.
The IBPF algorithm enables us to directly estimate the parameters of models describing high-dimensional partially-observed nonlinear dynamic systems via likelihood maximization.
The ability to directly estimate parameters of Model~3 is responsible for the large increase in model likelihoods reported in Table~\ref{tab:likes}.
Simulations from the fitted model are displayed in Fig.~\ref{fig:h3spatsims}.
\jwc{Should there be more added / said to this paragraph, to demonstrate the novelty of being able to fit a model like Model 3? Maybe add a paragraph between the first and second paragraphs of this sub-section.}

@@

\jwc{Should we make benchmark modeling a sub-section of model-diagnostics? I think this could (A) add some organization to the paper so that readers / reviewers can clearly see how benchmarks help answer the question we posed: what indications should we look for in data to assess trustworthiness of data (B) further emphasize the importance of benchmarks, which is something the reviewers seemed to have found useful. }
One common approach to assess a mechanistic model's goodness-of-fit is to compare simulations from the fitted model to the observed data.
Visual inspection may indicate defects in the model, or may  suggest that the observed data are a plausible realization of the fitted model.
While visual comparisons can be informative, they provide only a weak and informal measure of the goodness-of-fit of a model.
The study by Lee et al.~\cite{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were considerably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
Alternative forms of model validation should therefore be used in conjunction with visual comparisons of simulations to observed data.

>>>>

    u_cond_ll <- t(sapply(h3_bpf, function(x) rowSums(x@block.cond.loglik)))
    results$u_cond_ll <- apply(u_cond_ll, 2, logmeanexp, se = TRUE)

@@

\new{\editMechModels}

@@

\subsubsection*{Calibrating Model~3 Parameters}\label{sec:fit3}
Model~3 describes cholera dynamics in Haiti using a metapopulation model, where the hidden states in each administrative department has an effect on the dynamics in other departments.
The decision to address metapopulation dynamics using a spatially explicit model, rather than to aggregate over space, is double-edged.
Evidence for the former approach has been provided in previous studies \cite{king15}, including the specific case of heterogeneity between Haitian departments in cholera transmission \cite{collins14}.
However, a legitimate preference for simplicity can support a decision to consider nationally aggregated models \cite{saltelli20,green15}.

\new{\editMetaPop}

One issue that arises when fitting spatially explicit models is that parameter estimation techniques based on the particle filter become computationally intractable as the number of spatial units increases.
This is a result of the approximation error of particle filters growing exponentially in the dimension of the model \cite{rebeschini15,park20}.
To avoid the approximation error present in high-dimensional models, Lee et al.~\cite{lee20} simplified the problem of estimating the parameters of Model~3 by creating an approximate version of the model where the units are independent given the observed data.
Reducing a spatially coupled model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
Because the simplified, spatially-decoupled version of Model~3 implemented in \cite{lee20} relies on the observed cholera cases, the calibrated model cannot readily be used to obtain forecasts.
Therefore, in order to obtain model forecasts, Lee et al.~\cite{lee20} used the parameters estimates from the spatially-decoupled approximation of Model~3 to obtain forecasts using the fully coupled version of the model.
This approach of model calibration and forecasting avoids the issue of particle depletion, but may also be problematic.
One concern is that cholera dynamics in department $u$ are highly related to the dynamics in the remaining departments;
calibrating model parameters while conditioning on the observed cases in other departments may therefore lead to an over-dependence on observed cholera cases.
Another concern is that the two versions of the model are not the same, resulting in sub-optimal parameter estimates for the spatially coupled model,
as parameters that maximize the likelihood of the decoupled model almost certainly do not maximize the likelihood of the fully coupled model.
These two concerns may explain the unrealistic forecasts and low likelihood of Model~3 of Lee et al.~\cite{lee20} (see Table~\ref{tab:likes}).

At the time Lee et al.~\cite{lee20} conducted their study, there was no known algorithm that could readily be used to maximize the likelihood of an arbitrary meta-population POMP model with coupled spatial dynamics, which justifies the spatial decoupling approximation that was used to calibrate model parameters.
For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the \old{iterated block particle filter (IBPF)} \new{IBPF} algorithm \cite{ionides22}.
This algorithm extends the work of Ning and Ionides \cite{ning21ibpf}, who provided theoretic justification for the version of the algorithm that only estimates unit-specific parameters.
The IBPF algorithm enables us to directly estimate the parameters of models describing high-dimensional partially-observed nonlinear dynamic systems via likelihood maximization.
The ability to directly estimate parameters of Model~3 is responsible for the large increase in model likelihoods reported in Table~\ref{tab:likes}.
Simulations from the fitted model are displayed in Fig.~\ref{fig:h3spatsims}.

@@

One common approach to assess a mechanistic model's goodness-of-fit is to compare simulations from the fitted model to the observed data.
Visual inspection may indicate defects in the model, or may  suggest that the observed data are a plausible realization of the fitted model.
While visual comparisons can be informative, they provide only a weak and informal measure of the goodness-of-fit of a model.
The study by Lee et al.~\cite{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were considerably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
Alternative forms of model validation should therefore be used in conjunction with visual comparisons of simulations to observed data.
